\pkg{pksensi}: An R package to apply sensitivity analysis in pharmacokinetic models

### 2019-02-01 manuscript/manuscript.Rmd
update manuscript

Sensitivity analysis is a mathematical technique to investigate how variations in model parameters affect model outputs. An increasing number of studies use sensitivity analysis to determine which model parameters contribute to high variation in model predictions (Ferretti et al., 2016). Also, this technique has been applied in pharmacology and toxicology research (Loizou et al., 2015; McNally et al., 2012; Scherholz et al., 2018). Pharmacokinetic modeling describes the changes in the concentrations or amounts of a substance within model compartments over time. The goal of sensitivity analysis in pharmacokinetic research is to examine the sensitivity of output variables (e.g. compound concentration in blood or tissues) in pharmacokinetic models responds to input parameters, such as anatomical, physiological, and kinetic constants (McNally et al., 2011). It can be further applied to parameter prioritization and parameter fixing before model calibration (Hsieh et al., 2018).

In our previous work (Hsieh et al., 2018), we developed an approach to apply global sensitivity analysis workflow to reduce the computational burden in the Bayesian,  Markov Chain Monte Carlo (MCMC)-based calibration process of a physiologically based pharmacokinetic (PBPK) model. We used GNU MCSim (Bois, 2009), an effective simulation package for Bayesian population PBPK modeling, to calibrate the model. We found that the extended Fourier Amplitude Sensitivity Test (eFAST), a type of global sensitivity analysis algorithm, had the best balance of efficiency and accuracy for a complex, multi-compartment, multi-dataset, and multi-metabolite PBPK model. Also, we found some efficient visualization approaches that can be used to distinguish between “influential” and “non-influential” parameters. We also found a useful approach for communicating the parameter sensitivity in decision making.

Most of sensitivity analysis tools are constructed using scientific computing software, such as GNU Octave/Matlab (Pianosi et al., 2015), python (Herman and Usher, 2017), and R (Ghanem et al., 2017). Previously, we mainly used R with the sensitivity package (Pujol et al., 2017), which is a practical tool to conduct local and global sensitivity analysis. The sensitivity package includes some functions to generate the parameter sequences by using different computing algorithms. Also, it provides a feasible way to integrate external modeling results. This computational approach can effectively solve the heavy computing burden of numerical solutions within the pure R environment. Several R packages, such as FME, multisensi, and ODEsensitivity include sensitivity analysis tools and are freely available in the Comprehensive R Archive Network. The FME package contains a basic method for performing local sensitivity analysis for dynamic models (Soetaert and Petzoldt, 2016). The multisensi package provides sensitivity analysis for models with multivariate output (Bidot et al., 2018). ODEsensitivity can perform sensitivity analysis in ordinary differential equation (ODE) models (Weber et al., 2018).  It uses the ODE interface from the deSolve package, connecting it with the sensitivity analysis from the sensitivity package, and can perform sensitivity analysis on models with multivariate output. Although these tools provide various approaches to perform sensitivity analysis, we have not find a suitable package that provides functions with which users can visualize and distinguish the “non-influential” model parameters and can further apply to parameter fixing in pharmacokinetic modeling.

Here we develop an R package, called pksensi, which is designed to make sensitivity analysis more accessible and reproducible in pharmacological and toxicological researches. This package can investigate both parameter uncertainty and sensitivity in pharmacokinetic models, such as PBPK, and advanced compartment absorption and transit models with multivariate model output. The design concepts of pksensi are:

*	Cross-platform: Models can run on Windows/MacOS/Linux
*	Freely available: All related packages are free and open source
*	Integrated application: Users can run pharmacokinetic models in R with script that were written in C or GNU MCSim 
*	Decision making: The output results and visualization tools can be used to easily determine which parameters have “non-influential” effects on the model output and can be fixed in model calibration.

@@

We adopted eFAST (extended Fourier Amplitude Sensitivity Test), a widely used global sensitivity analysis approach in biomathematical modeling, especially for ordinary differential equation-based dynamical models, in this package (Saltelli et al., 1999). Most of the available eFAST functions, such as the algorithm to generate the parameter space and to set the sampling frequency, were sourced from the R \pkg{sensitivity} package (Pujol et al., 2017). To test the convergence and robustness of the sensitivity measurement, we included a random phase-shift approach to replicate sampling from random starting points across parameter space. This can be written as:

@@

where $x_i$ is the nominal value of the i-th parameter, and $\omega_i$  is a vector giving the set of frequencies, one frequency for each parameter. The default set of frequencies is based on the study from Saltelli et al. (1999). The first frequency of the vector $\omega$ is assigned to each factor $x_i$ in turn (corresponding to the estimation of Sobol indices of first and total order, respectively). Through the random phase-shift, the robust result of the sensitivity measurement should be similar across replications under the same sample size.

To investigate the convergence of sensitivity indices, we adopted the exam approach proposed by Sarrazin et al. (2016). This method quantitatively assesses convergence by computing the width of 95% confidence intervals for all parameters across all time-points and output variables. To integrate the random phase-shift with eFAST, we built the \proglang{rfast99} function, which can be used to perform the sensitivity test for the whole pharmacokinetic modeling process. The function \proglang{rfast99} in pksensi can sample and generate the testing parameter matrix based on a given argument of sample size n and probability distribution (e.g., mean/s.d. of normal distribution, meanlog/sdlog of log-normal distribution, and minimum/maximum of uniform distribution). The number of model evaluations is equal to the sample size times the number of model parameters.

@@

The \pkg{pksensi} provides a useful function to solve ordinary differential equations in pharmacokinetic models using analytical and numerical approaches. The solution can be performed under the pure R coding environment by linking pksensi with the deSolve package (Soetaert et al., 2018). However, the most efficient way to achieve high computational speed is through compiled, lower-level languages, such as FORTRAN, C, or C++. The \pkg{pksensi} includes a function to compile and create dynamic-link libraries (.dll) on Windows and shared objects (.so) on Unix-liked systems (e.g., Linux and MacOS). This compiled file can load and execute through the build-in function in R. However, Windows users need to install Rtools or MinGW to compile the source code by using the GNU GCC compiler. More requirements are detailed in the study from Soetaert et al. (2010).  The C implementation of the example model can be found as an example within MCSim v6.0.1 (Bois, 2009). The \pkg{pksensi} can also link with MCSim to compile the model code, used in solving each system of equations.

@@

The output of sensitivity analysis returns a list that contains the given time-points and values of all state variables (e.g., chemical concentrations in blood). This format is particularly suited for further graphical routines within pkseni. Also, a plot can can create time-dependent sensitivity measurements of first and total order effect for all parameters in one figure. Furthermore, pksensi includes functions to check the convergence and sensitivity of model parameters, providing a means to assess the robustness of the sensitivity measurement. We also developed a “cut-off”-based approach in this package to accurately distinguish between “influential” and “non-influential” parameters. Finally, pksensi includes numerous visualization tools for the effective investigation and communication of results in decision making.

@@

## One-compartment taxicokinetic model

```{r setup, include = FALSE}
library(pksensi)
#mcsim_install()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height=4, 
  fig.width=6
)

fn = local({ # not used function for fig caption
  i = 0
  function(x) {
    i <<- i + 1
    paste('Figure ', i, ': ', x, sep = '')
  }
})

```

@@

In this example, We use a simple, one-compartment PK model from `httk` package [@JSSv079i04] to demonstrate how `pksensi` can be applied to pharmacokinetic studies. The differential equations for the one-compartment pharmacokinetic model can be written as:

@@

In the beginning, we need to pre-install GNU MCSim [@JSSv002i09] and related compiler to let us generate .c file and executable file. The GNU MCSim can be installed by following the instruction in GNU MCSim’s manual on https://www.gnu.org/software/mcsim/mcsim.html or using the build-in function `mcsim_install()` in `pksensi`. The GNU compiler is necessary for users that use Linux or MacOS. For Windows users, you should install Rtools on https://cran.r-project.org/bin/windows/Rtools/ and use `Sys.setenv()` to set the working path of compiler. The  `Sys.which("gcc")` and `system('g++ -v')` can check whether we can run compiler correctly.

We first implemented this model in R by compiling the file written in C. pksensi allows users to select the preferred method to solve the pharmacokinetic model, either with the `deSolve` package or with GNU MCSim through the compile function. However, running model under GNU MCSim native code can have faster speed to obtain the model outputs.

The following R script can download and compile the model description file (`pbtk1cpt.model`) and use `deSolve` package [@JSSv033i09] to solve ordinary differential equations in our model. The example model code of one-compartment PBTK model is available with pksensi package:

@@

Then, use `compile_model()` to generate the executable files (`pbtk1cpt.dll` on Windows or `pbtk1cpt.so` on other systems) and R file (`pbtk1cpt_inits.R`) with default input parameters and initial state settings with the definition of `application = "R"`.

@@

In the current setting, we assumed the initial condition of the intake chemical to be 10 mol. The `initParms` and `initStates` functions were used to customize the parameter values and the initial state that will be used in the `solve_fun` function. These parameter value can be adopted from the `httk` package, which includes physico-chemical and drug biological properties for 553 chemicals. In this case, we used the parameter value of theophylline in this example. The given `vdist`, `ke`, and `kgutabs` are 0.74, 0.28, and 2.18, respectively.

Through `ode` function in `deSolve` package, we can visualize the pharmacokinetic according to the given parameter conditions such as time points (`times`):

```{r, eval=T}
times <- seq(from = 0.01, to = 24.01, by = 1)
y <- deSolve::ode(initState, times, func = "derivs", parms = parms, 
                  dllname = mName, initfunc = "initmod", nout = 1, outnames = Outputs)
```

@@

To conduct sensitivity analysis for the parameters in one-compartment pharmacokinetic model in this case, we want to quantify the impact of these three parameters on the chemical concentration in plasma during 24-hour time period post intake. We assume a uniform distribution for the estimate for each parameter with the coefficient of uncertainty within 50%. The parameter ranges are assumed to be (0.37, 1.12) for `vdist`, (0.0058, 0.0174) for `ke`, and (0.045, 0.136) for `kgutabs`. The sample number determines the robustness of the result of sensitivity analysis. Higher sample numbers can generate narrow confidence intervals for sensitivity measurements across different replications. However, they might cause heavy computational burden for complex models. Here we use a sample number of 400 with 20 replications:

@@

To create the time-dependent sensitivity measurement, we set the time duration from 0.01 to 24.01 hours in the example. The `initParmsfun` is used to generate the sampling value for each parameter. The `outnames`, `dllname`, `func`, `initfunc` are based on the arguments from the ode function in `deSolve` package. The details of the model structure and these arguments are defined in `pbtk1comp.c`. and `pbtk1comp_inits.R`. Finally, the `tell2` function is used to integrate the parameter values and the output results of numerical analysis that were generated and stored in variables x and y. The result of object x is an object of rfast99, which has specific `print`, `plot`, and `check` method. The print function gives the sensitivity and convergence indices for main, interaction, and total order at each time point. In addition to print out the result of sensitivity analysis, the more efficient way to distinguish the influence of model parameter is to visualize them. The time-dependent sensitivity indices are shown in Figure 2.

```{r, eval=T, fig.cap = 'Figure 2. Time-dependent sensitivity indices of the plasma concentration estimated from one-compartment PBTK model during 24 hour time period intake.'}
plot(x)
```

Here, we can find that `vdist` and `ke` are dominating the plasma concentration in the before and after 5-hour post chemical intake, respectively, representing that the elimination is a key parameter to dominate the plasma concentration. Besides, the `kgutabs` only plays a crucial role to determine the plasma concentration in the first hour. The relationship between concentration and the parameters can be plotted as follow (Figure 3):

@@

The x is a list of class "rfast99", containing all the input arguments detailed before and the calculated sensitivity indices of first order (`mSI`), interaction (`iSI`), and total order (`tSI`). The convergence indices are also stored in the list named `mCI`, `iCI`, and `tCI`. The parameter values are stored in an array `x$a` with c(model evaluation, replication, parameters).

@@

The `check()` is a useful function to determine which parameters have relative lower sensitivity measurement across the given time interval, and therefore can be applied parameter fixing in model calibration. The argument of `SI.cutoff` is setting at 0.5 to detect the relative non-influential parameters in this case.

@@

Based on the sensitivity measurement of the total order, the result shows that `kgutabs` has relative lower measurement of sensitivity index.

@@

Alternatively, to solve ODE by using GNU MCSim, we need to change the argument to `application = mcsim` in `compile_model()`. Rather than apply R `deSolve` to solve differential equations, the GNU MCSim can provide higher computational speed in global sensitivity analysis.

@@

Sililiar to `solve_fun` that can define the initial parameter and state values through input function, the `solve_mcsim` has a `condition` argument that is used to givien the specific input value such as oral dose or fixing parameter value or initial state variable.

@@

Under the same given condition, it takes 5-6 (`deSolve`) and 1-2 (GNU MCSim) seconds to solve model. The `solve_mcsim()` shows the better computational performance than `solve_fun()` in `pksensi`.

@@

The aim of this section is to reproduce our previous published [@fphar201800588] result of global sensitivity analysis for acetaminophen PBPK model through `pksensi`. The model codes are included in this package and can be generated through `pbpk_apap_model()`. We applied the global sensitivity analysis workflow to the original published model with 21 model parameters [@s13318-015-0253-x]. The descriptions of each parameter and the sampling ranges are list in Table 1.

<Table 1>

@@

Generally, The wide range of parameter value might cause the computational error in the solver. One of the effective ways to prevent this problem is to adjust the value of relative and absolute error tolerance to control the error appearance by resetting these parameters in a lower value. The `generate_infile()` provide the arguments of `rtol` and `atol` that can be adjusted to prevent the unwanted error. However, the modification will slow down the computational speed. Therefore, the alternative method to prevent the computational error is to detect the crucial parameter range that causes the problem. Also, setting the maximum number of (internally defined) steps to higher value instead of using the default value (500) can prevent this problem. The maximum number of step is set to 5000 in this case.

In this test case, we adjusted the range of `SULT_VmaxC` and `UGT_VmaxC` from U(0, 15) to U(0, 10). The relative and absolute error tolerance were set to 1e-7 and 1e-9, respectively, to prevent the computational error in MCSim,

@@

After creating the `pbpk_apap.model` in the working directory, the next step is to generate the executable files (`mcsim.pbpk_apap`) through `compile_model()`.

@@

To improve the computational speed, this case only uses MCSim to estimate the concentration of acetaminophen (APAP) and its metabolites glucuronide (AG) and sulfate (AS) in plasma. The setting oral dose of APAP is 20 mg/kg in this example. Generally, the input dosing method can be defined through the `condition` argument. Since the unit of the given dose is mg/kg, the `mgkg_flag` is set to 1 to declare the statement. More definition of input can be found in the section of input functions in GNU MCSim User’s Manual (https://www.gnu.org/software/mcsim/mcsim.html#Input-functions).

@@

The plotting function can output the result of time-dependent sensitivity measurement to determine the parameter impact on model output over time (Figure 1).

```{r, eval=T, fig.height=8, fig.width=8, fig.cap = 'Figure 4. '}
plot(x, vars = "lnCPL_AG_mcgL")
```

The uncertainty analysis is a crucial step before model calibration. We can apply uncertainty analysis through the `pksim()` by the given name of output variables (Figure 2). Through this visualization approach, we can recognize whether the simulated outputs can accurately simulate the same concentration profile as the in-vivo experiment under the setting of parameter ranges. In Figure 2, all experiment points are included in the intervals, representing the acceptable set of the parameter range.

>>>>

Sensitivity analysis is a mathematical technique to investigate how variations in model parameters affect model outputs. An increasing number of studies use sensitivity analysis to determine which model parameters contribute to high variation in model predictions \citep{ferretti2016trends}. Also, this technique has been applied in pharmacology and toxicology research \citep{loizou2015application, mcnally2012reconstruction}. Pharmacokinetic modeling describes the changes in the concentrations or amounts of a substance within model compartments over time. The goal of sensitivity analysis in pharmacokinetic research is to examine the sensitivity of output variables (e.g. compound concentration in blood or tissues) in pharmacokinetic models responds to input parameters, such as anatomical, physiological, and kinetic constants \citep{mcnally2011workflow}. It can be further applied to parameter prioritization and parameter fixing before model calibration \citep{fphar201800588}.

In our previous work \citep{fphar201800588}, we developed an approach to apply global sensitivity analysis workflow to reduce the computational burden in the Bayesian,  Markov Chain Monte Carlo (MCMC)-based calibration process of a physiologically based pharmacokinetic (PBPK) model. We used GNU MCSim \citep{JSSv002i09}, an effective simulation package for Bayesian population PBPK modeling, to calibrate the model. We found that the extended Fourier Amplitude Sensitivity Test (eFAST), a type of global sensitivity analysis algorithm, had the best balance of efficiency and accuracy for a complex, multi-compartment, multi-dataset, and multi-metabolite PBPK model. Also, we found some efficient visualization approaches that can be used to distinguish between “influential” and “non-influential” parameters. We also found a useful approach for communicating the parameter sensitivity in decision making.

Most of sensitivity analysis tools are constructed using scientific computing software, such as \proglang{GNU Octave}/\proglang{Matlab} \citep{pianosi2015matlab}, \proglang{python} \citep{herman2017salib}, and \proglang{R} \citep{iooss2017introduction}. Previously, we mainly used \proglang{R} with the \pkg{sensitivity} package \citep{R-sensitivity}, which is a practical tool to conduct local and global sensitivity analysis. The sensitivity package includes some functions to generate the parameter sequences by using different computing algorithms. Also, it provides a feasible way to integrate external modeling results. This computational approach can effectively solve the heavy computing burden of numerical solutions within the pure \proglang{R} environment. Several \proglang{R} packages, such as \pkg{FME}, \pkg{multisensi}, and \pkg{ODEsensitivity} include sensitivity analysis tools and are freely available in the Comprehensive R Archive Network (CRAN). The \pkg{FME} package contains a basic method for performing local sensitivity analysis for dynamic models \citep{JSSv033i03}. The \pkg{multisensi} package provides sensitivity analysis for models with multivariate output \citep{R-multisensi}. \pkg{ODEsensitivity} can perform sensitivity analysis in ordinary differential equation (ODE) models \citep{R-ODEsensitivity}. It uses the ODE interface from the deSolve package, connecting it with the sensitivity analysis from the sensitivity package, and can perform sensitivity analysis on models with multivariate output. Other related packages such as \pkg{mtk} \citep{RJ-2015-031} and \pkg{fast} \citep{R-fast} are designed to invetigate the influence of model parameter. Although these tools provide various approaches to perform sensitivity analysis, we have not find a suitable package that provides functions with which users can visualize and distinguish the “non-influential” model parameters and can further apply to parameter fixing in pharmacokinetic modeling.  
Here we develop an \proglang{R} package, called \pkg{pksensi}, which is designed to make sensitivity analysis more accessible and reproducible in pharmacological and toxicological researches. This package can investigate both parameter uncertainty and sensitivity in pharmacokinetic models, such as PBPK, and advanced compartment absorption and transit models with multivariate model output. The design concepts of \pkg{pksensi} are:

*	Cross-platform: Models can run on Windows/MacOS/Linux
*	Freely available: All related packages are free and open source
*	Integrated application: Users can run pharmacokinetic models in \proglang{R} with script that were written in \proglang{C} or \proglang{GNU MCSim} 
*	Decision making: The output results and visualization tools can be used to easily determine which parameters have "non-influential" effects on the model output and can be fixed in model calibration.

@@

We adopted eFAST (extended Fourier Amplitude Sensitivity Test), a widely used global sensitivity analysis approach in biomathematical modeling, especially for ordinary differential equation-based dynamical models, in this package \citep{saltelli1999quantitative}. Most of the available eFAST functions, such as the algorithm to generate the parameter space and to set the sampling frequency, were sourced from the \proglang{R} \pkg{sensitivity} package \citep{R-sensitivity}. To test the convergence and robustness of the sensitivity measurement, we included a random phase-shift approach to replicate sampling from random starting points across parameter space. This can be written as:

@@

where $x_i$ is the nominal value of the i-th parameter, and $\omega_i$  is a vector giving the set of frequencies, one frequency for each parameter. The default set of frequencies is based on the study from \cite{saltelli1999quantitative}. The first frequency of the vector $\omega$ is assigned to each factor $x_i$ in turn (corresponding to the estimation of Sobol indices of first and total order, respectively). Through the random phase-shift, the robust result of the sensitivity measurement should be similar across replications under the same sample size.

To investigate the convergence of sensitivity indices, we adopted the exam approach proposed by \citep{sarrazin2016global}. This method quantitatively assesses convergence by computing the width of 95% confidence intervals for all parameters across all time-points and output variables. To integrate the random phase-shift with eFAST, we built the \code{rfast99} function, which can be used to perform the sensitivity test for the whole pharmacokinetic modeling process. The function \code{rfast99} in \pkg{pksensi} can sample and generate the testing parameter matrix based on a given argument of sample size n and probability distribution (e.g., mean/s.d. of normal distribution, meanlog/sdlog of log-normal distribution, and minimum/maximum of uniform distribution). The number of model evaluations is equal to the sample size times the number of model parameters.

@@

The \pkg{pksensi} provides a useful function to solve ordinary differential equations in pharmacokinetic models using analytical and numerical approaches. The solution can be performed under the pure \proglang{R} programming environment by linking pksensi with the \pkg{deSolve} package \citep{JSSv033i09}. However, the most efficient way to achieve high computational speed is through compiled, lower-level languages, such as \proglang{FORTRAN}, \proglang{C}, or \proglang{C++}. The \pkg{pksensi} includes a function to compile and create dynamic-link libraries (.dll) on Windows and shared objects (.so) on Unix-liked systems (e.g., Linux and MacOS). This compiled file can load and execute through the build-in function in \proglang{R}. However, Windows users need to install Rtools or MinGW to compile the source code by using the GNU GCC compiler. More requirements are detailed in the study from \citep{soetaertr}.  The \proglang{C} implementation of the example model can be found as an example within \proglang{GNU MCSim} v6.0.1 \citep{JSSv002i09}. The \pkg{pksensi} can also link with \proglang{GNU MCSim} to compile the model code, used in solving each system of equations.

@@

The output of sensitivity analysis returns a list that contains the given time-points and values of all state variables (e.g., chemical concentrations in blood). This format is particularly suited for further graphical routines within \pkg{pksensi}. Also, a plot can can create time-dependent sensitivity measurements of first and total order effect for all parameters in one figure. Furthermore, \pkg{pksensi} includes functions to check the convergence and sensitivity of model parameters, providing a means to assess the robustness of the sensitivity measurement. We also developed a "cut-off"-based approach in this package to accurately distinguish between "influential" and "non-influential" parameters. Finally, \pkg{pksensi} includes numerous visualization tools for the effective investigation and communication of results in decision making.

@@

## One-compartment toxicokinetic model

```{r setup, include = FALSE}
library(pksensi)
options(xtable.comment = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height=4, 
  fig.width=6
)
```

@@

In this example, We use a simple, one-compartment PK model from \pkg{httk} package \citep{JSSv079i04} to demonstrate how \pkg{pksensi} can be applied to pharmacokinetic studies. The differential equations for the one-compartment pharmacokinetic model can be written as:

@@

In the beginning, we need to pre-install \pkg{GNU MCSim} \citep{JSSv002i09} and related compiler to let us generate .c file and executable file. The \pkg{GNU MCSim} can be installed by following the instruction in \pkg{GNU MCSim}’s manual on https://www.gnu.org/software/mcsim/mcsim.html or using the build-in function \code{mcsim_install()} in \pkg{pksensi}. The GNU compiler is necessary for users that use Linux or MacOS. For Windows users, you should install Rtools on https://cran.r-project.org/bin/windows/Rtools/ and use \code{Sys.setenv()} to set the working path of compiler. In \proglang{R} environment, The  \code{Sys.which("gcc")} and \code{system('g++ -v')} can check whether we can run compiler correctly.

We first implemented this model in \proglang{R} by compiling the file written in \proglang{C}. \pkg{pksensi} allows users to select the preferred method to solve the pharmacokinetic model, either with the pkg{deSolve} package or with \pkg{GNU MCSim} through the compile function. However, running model under \pkg{GNU MCSim} native code can have faster speed to obtain the model outputs.

The following \proglang{R} script can download and compile the model description file \code{pbtk1cpt.model} and use code{ode} function in \pkg{deSolve} package \citep{JSSv033i09} to solve ordinary differential equations in our model. The example model code of one-compartment PBTK model is available with \pkg{pksensi} package:

@@

Then, use \code{compile_model()} to generate the executable files (\code{pbtk1cpt.dll} on Windows or \code{pbtk1cpt.so} on other systems) and \proglang{R} file (\code{pbtk1cpt_inits.R}) with default input parameters and initial state settings with the definition of \code{application = "R"}.

@@

In the current setting, we assumed the initial condition of the intake chemical to be 10 mol. The \code{initParms} and \code{initStates} functions were used to customize the parameter values and the initial state that will be used in the \code{solve_fun()} function. These parameter value can be adopted from the \pkg{httk} package, which includes physico-chemical and drug biological properties for over 500 chemicals. In this case, we used the parameter value of theophylline in this example. The given \code{vdist}, \code{ke}, and \code{kgutabs} are 0.74, 0.28, and 2.18, respectively.

Through \code{ode} function in \pkg{deSolve} package, we can visualize the pharmacokinetic according to the given parameter conditions such as time points (\code{times}):

```{r, eval=T}
library(deSolve)
times <- seq(from = 0.01, to = 24.01, by = 1)
y <- ode(initState, times, func = "derivs", parms = parms, 
         dllname = mName, initfunc = "initmod", nout = 1, outnames = Outputs)
```

@@

To conduct sensitivity analysis for the parameters in one-compartment pharmacokinetic model in this case, we want to quantify the impact of these three parameters on the chemical concentration in plasma during 24-hour time period post intake. We assume a uniform distribution for the estimate for each parameter with the coefficient of uncertainty within 50%. The parameter ranges are assumed to be (0.37, 1.12) for \code{vdist}, (0.0058, 0.0174) for \code{ke}, and (0.045, 0.136) for \code{kgutabs}. The sample number determines the robustness of the result of sensitivity analysis. Higher sample numbers can generate narrow confidence intervals for sensitivity measurements across different replications. However, they might cause heavy computational burden for complex models. Here we use a sample number of 400 with 20 replications:

@@

To create the time-dependent sensitivity measurement, we set the time duration from 0.01 to 24.01 hours in the example. The \code{initParmsfun} is used to generate the sampling value for each parameter. The \code{outnames}, \code{dllname}, \code{func}, \code{initfunc} are based on the arguments from the ode function in \pkg{deSolve} package. The details of the model structure and these arguments are defined in \code{pbtk1comp.c}. and \code{pbtk1comp_inits.R}. Finally, the \code{tell2()} function is used to integrate the parameter values and the output results of numerical analysis that were generated and stored in variables x and y. The result of object x is an object of rfast99, which has specific \code{print()}, \code{plot()}, and \code{check()} method. The print function gives the sensitivity and convergence indices for main, interaction, and total order at each time point. In addition to print out the result of sensitivity analysis, the more efficient way to distinguish the influence of model parameter is to visualize them. The time-dependent sensitivity indices are shown in Figure 2.

```{r, eval=T, fig.cap = 'Figure 2. Time-dependent sensitivity indices of the plasma concentration estimated from one-compartment PBTK model during 24-hour time period intake.'}
plot(x)
```

Here, we can find that \code{vdist} and \code{ke} are dominating the plasma concentration in the before and after 5-hour post chemical intake, respectively, representing that the elimination is a key parameter to dominate the plasma concentration. Besides, the \code{kgutabs} only plays a crucial role to determine the plasma concentration in the first hour. The relationship between concentration and the parameters can be plotted as follow (Figure 3):

@@

The \code{x} is a list of class \code{rfast99}, containing all the input arguments detailed before and the calculated sensitivity indices of first order (\code{mSI}), interaction (\code{iSI}), and total order (\code{tSI}). The convergence indices are also stored in the list named \code{mCI}, \code{iCI}, and \code{tCI}. The parameter values are stored in an array \code{x$a} with c(model evaluation, replication, parameters).

@@

The \code{check()} is a useful function to determine which parameters have relative lower sensitivity measurement across the given time interval, and therefore can be applied parameter fixing in model calibration. The argument of \code{SI.cutoff} is setting at 0.5 to detect the relative non-influential parameters in this case.

@@

Based on the sensitivity measurement of the total order, the result shows that \code{kgutabs} has relative lower measurement of sensitivity index.

@@

Alternatively, to solve ODE by using \pkg{GNU MCSim}, we need to change the argument to \code{application = mcsim} in \code{compile_model()}. Rather than apply \pkg{deSolve} to solve differential equations, the \pkg{GNU MCSim} can provide higher computational speed in global sensitivity analysis.

@@

Similar to \code{solve_fun()} function that can define the initial parameter and state values through input function, the \code{solve_mcsim()} has a `condition` argument that is used to givien the specific input value such as oral dose or fixing parameter value or initial state variable.

@@

Under the same given condition, it takes 5-6 seconds (\pkg{deSolve}) and 1-2 seconds (\pkg{GNU MCSim})  to solve model. The \code{solve_mcsim()} shows the better computational performance than \code{solve_fun()} in \pkg{pksensi}.

@@

The aim of this section is to reproduce our previous published result \citep{fphar201800588} of global sensitivity analysis for acetaminophen PBPK model through \pkg{pksensi}. The model codes are included in this package and can be generated through \code{pbpk_apap_model()}. We applied the global sensitivity analysis workflow to the original published model with 21 model parameters [@s13318-015-0253-x]. The descriptions of each parameter and the sampling ranges are list in Table 1.

```{r, results='asis',message=FALSE,warning=FALSE, echo = F}
library(xtable)

#Nominal value
Tg <- log(0.23)
Tp <- log(0.033)
CYP_Km <- log(130)
SULT_Km_apap <- log(300)
SULT_Ki <- log(526)
SULT_Km_paps <- log(0.5)
UGT_Km <- log(6.0e3)
UGT_Ki <- log(5.8e4)
UGT_Km_GA <-log(0.5)
Km_AG <- log(1.99e4)
Km_AS <- log(2.29e4)

r <- 1.96 # exp(1.96)/exp(-1.96) ~ 50

x <- c("Tg", "Tp", "CYP\\_Km", "CYP\\_VmaxC",
       "SULT\\_Km\\_apap","SULT\\_Ki","SULT\\_Km\\_paps","SULT\\_VmaxC",
       "UGT\\_Km","UGT\\_Ki","UGT\\_Km\\_GA","UGT\\_VmaxC",
       "Km\\_AG","Vmax\\_AG","Km\\_AS","Vmax\\_AS",
       "kGA\\_syn","PAPS\\_syn", "CLC\\_APAP","CLC\\_AG","CLC\\_AS"
       )

y <- c("Gatric emptying time constant",
       "GI perfusion time constant",
       "Cytochrome P450 metabolism, Km",
       "Cytochrome P450 metabolism, VMax",
       "Sulfation pathway acetaminophen, Km",
       "Sulfation pathway substrate inhibition, Ki",
       "Sulfation pathway PAPS, Km",
       "Sulfation pathway acetaminophen, Vmax",
       "Glucronidation pathway acetaminophen, Km",
       "Glucronidation pathway substrate inhibition, Ki",
       "Glucronidation pathway GA, Km",
       "Glucronidation pathway acetaminophen, Vmax",
       "APAP-G hepatic transporter, Km",
       "APAP-G hepatic transporter, Vmax",
       "APAP-S hepatic transporter, Km",
       "APAP-S hepatic transporter, Vmax",
       "UDPGA synthesis",
       "PAPS synthesis",
       "APAP clearance",
       "APAP-G clearance",
       "APAP-S clearance"
       )

z <- c("$h$", "$h$", "$\\mu{M}$", "$\\mu{mole}/h\\cdot{BW}^{0.75}$",
       "$\\mu{M}$", "$\\mu{M}$", "$-$", "$\\mu{mole}/h\\cdot{BW}^{0.75}$",
       "$\\mu{M}$", "$\\mu{M}$", "$-$", "$\\mu{mole}/h\\cdot{BW}^{0.75}$",
       "$\\mu{M}$", "$\\mu{mole}/h$", "$\\mu{M}$", "$\\mu{mole}/h$",
       "$1/h$", "$1/h$",
       "$L/h\\cdot{BW}^{0.75}$", "$L/h\\cdot{BW}^{0.75}$", "$L/h\\cdot{BW}^{0.75}$"
       )

min <- c(round(Tg-r, 3), round(Tp-r, 3), round(CYP_Km-r), round(log(0.14), 3),
         round(SULT_Km_apap-r, 3), round(SULT_Ki-r, 3), round(SULT_Km_paps-r), log(1),
         round(UGT_Km-r, 3), round(UGT_Ki-r, 3), round(UGT_Km_GA-r), log(1),
         round(Km_AG-r, 3), round(log(1.09e3), 3), round(Km_AS-r), round(log(1.09e3), 3),
         log(1), log(1), round(log(2.48e-3), 3), round(log(2.48e-3), 3), round(log(2.48e-3), 3)
         )

max <- c(round(Tg+r, 3), round(Tp+r, 3), round(CYP_Km+r), round(log(2900), 3),
         round(SULT_Km_apap+r, 3), round(SULT_Ki+r, 3), round(SULT_Km_paps+r), round(log(22026), 3),
         round(UGT_Km+r, 3), round(UGT_Ki+r, 3), round(UGT_Km_GA+r), round(log(22026), 3),
         round(Km_AG+r, 3), round(log(3.26e6), 3), round(Km_AS+r), round(log(3.26e6), 3),
         round(log(4.43e5), 3), round(log(4.43e5),3), 
         round(log(2.718), 3), round(log(2.718), 3), round(log(2.718), 3)
         )

df <- data.frame(x, y, z, min, max)
names(df) <- c("Parameter","Description", "Unit", "Min", "Max")

print(xtable(df, caption = "Description of sampling range of model parameter"),
      type = "latex", 
      caption.placement = "top", include.rownames = FALSE,
      sanitize.text.function = function(x) {x})
```

@@

Generally, The wide range of parameter value might cause the computational error in the solver. One of the effective ways to prevent this problem is to adjust the value of relative and absolute error tolerance to control the error appearance by resetting these parameters in a lower value. The \code{generate_infile()} provide the arguments of \code{rtol} and \code{atol} that can be adjusted to prevent the unwanted error. However, the modification will slow down the computational speed. Therefore, the alternative method to prevent the computational error is to detect the crucial parameter range that causes the problem. Also, setting the maximum number of (internally defined) steps to higher value instead of using the default value (500) can prevent this problem. The maximum number of step is set to 5000 in this case.

In this test case, we adjusted the range of \code{SULT_VmaxC} and \code{UGT_VmaxC} from U(0, 15) to U(0, 10). The relative and absolute error tolerance were set to 1e-7 and 1e-9, respectively, to prevent the computational error in \pkg{GNU MCSim},

@@

After creating the \code{pbpk_apap.model} file in the working directory, the next step is to generate the executable files (\code{mcsim.pbpk_apap}) through \code{compile_model()}.

@@

To improve the computational speed, this case only uses \pkg{GNU MCSim} to estimate the concentration of acetaminophen (APAP) and its metabolites glucuronide (AG) and sulfate (AS) in plasma. The setting oral dose of APAP is 20 mg/kg in this example. Generally, the input dosing method can be defined through the \code{condition} argument. Since the unit of the given dose is mg/kg, the \code{mgkg_flag} is set to 1 to declare the statement. More definition of input can be found in the section of input functions in \pkg{GNU MCSim} User’s Manual (https://www.gnu.org/software/mcsim/mcsim.html#Input-functions).

@@

The plotting function can output the result of time-dependent sensitivity measurement to determine the parameter impact on model output over time (Figure 4).

```{r, eval=T, fig.height=8, fig.width=8, fig.cap = ' '}
plot(x, vars = "lnCPL_AG_mcgL")
```

The uncertainty analysis is a crucial step before model calibration. We can apply uncertainty analysis through the `pksim()` by the given name of output variables (Figure 5). Through this visualization approach, we can recognize whether the simulated outputs can accurately simulate the same concentration profile as the in-vivo experiment under the setting of parameter ranges. In Figure 5, all experiment points are included in the intervals, representing the acceptable set of the parameter range.
